Phenotypic evolution is less constrained in complex food webs

### 2018-10-08 manuscript/manuscript.tex
update of current analyses

\documentclass[11pt]{article}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{graphicx} % for organizing figures
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

@@

The fitness landscape provides a unifying framework for linking the ecology and evolution of populations (Lande 2007; McPeek 2017). The average fitness of a population is a common currency in ecology and evolution, but usually goes by different names in each field. Ecologists refer to it as per-capita population growth rate ($dN/Ndt$), whereas evolutionary biologists call it the natural log of population mean fitness ($ln(\bar W_N)$). In addition to having different names, ecologists and evolutionary biologists have typically focused on different processes that shape the fitness landscape. For example, population ecologists have long studied the effect of a population's density on its per-capita growth rate (i.e. density-dependence, CITE Foundational and current work). In contrast, evolutionary biologists have focused on how the mean trait value of a population influences its average fitness, as this describes the direction and magnitude of natural selection (CITE foundational and current work). Therefore, the fitness landscape describes the joint ecological and evolutionary dynamics of a population in a given environment.

Community ecologists have extended the ecological side of the fitness landscape by incorporating network theory. Species-interaction networks, such as a food web describing who eats whom, provide an explicit representation of the biotic environment as they describe the interdependency of populations within an ecological community. This has provided an effective framework for predicting how changes in the biotic environment (e.g. density of directly and indirectly connected species) will impact population dynamics within species-rich communities. At the same time, evolutionary biologists have long recognized that changes in the biotic environment can alter the dynamics of natural selection. However, the biotic environment in which populations are evolving often remains a bit of a "black box" that's labelled by a general ecological process such as competition, predation, or mutualism. Because of this, it remains difficult to predict how changes in the biotic environment will affect the direction and magnitude of natural selection. Such predictions are urgently needed given the rapid changes in the biotic environment that most populations are currently experiencing throughout the world.

Here, we integrate species-interaction networks and the fitness landscape to empirically test how changes in the biotic environment -- network of species interactions -- affect the dynamics of natural selection. Specifically, we conducted a field experiment that manipulated the diversity of insect parasitoids that were able to impose selection on an abundant insect herbivore (*Iteomyia salicisverruca*)(Fig. 1). The larva of this herbivore species induce tooth-shaped galls when they feed on the developing leaves of willow trees (*Salix* sp., @Russo2006). These galls provide protection from generalist predators (e.g. ants, spiders), thus the network of interacting parasitoids provides a realistic representation of the biotic environment this insect herbivore is experiencing. Therefore, our manipulation of parasitoid diversity alters the diversity of interactions, or food-web complexity, that this insect herbivore experiences.

@@

@@

We conducted our study within a four-year old common garden of coastal willow (*Salix hookeriana*) located at Humboldt Bay National Wildlife Refuge (HBNWR) (40&deg;40'53"N, 124&deg;12'4"W) near Loleta, California, USA. This common garden consists of 26 different willow genotypes that were collected from a single population of willows growing around Humboldt Bay. Stem cuttings of each genotype (25 replicates per genotypes) were planted in a completely randomized design in two hectares of a former cattle pasture at HBNWR. Willows in our garden begin flowering in February and reach their peak growth in early August. During this study, willows had reached 5 - 9m in height. Further details on the genotyping and planting of the common garden are available in @Barbour2015.

@@

We setup our food-web manipulation across 128 plants soon after galls began developing on *S. hookeriana* in early June of 2013. 
These 128 plants came from eight different plant genotypes, spanning the range of trait variation observed in this willow population (@Barbour2015). 
On treatment plants (8 replicates per genotype), we enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species that attack during larva development (hereafter larval parasitoids). 
This treatment did not exclude the egg parasitoid *Platygaster* sp. which attacks prior to gall initiation (note that in Cecidomyiid midges, larva initiate gall development CITE). 
On control plants (8 replicates per genotype), we used flagging tape to mark 14 galled leaves per plant (~30 larva), allowing the full suite of parasitoids to attack *Iteomyia*. 
Marking galls with flagging tape ensured that we compared control and treatment galls with similar phenology when we collected galls later in the season. 
Our food-web manipulation altered the average number of trophic interactions that *Iteomyia* was exposed to from BLANK on control plants to BLANK on treatment plants. 
Thus, we refer to galls on control plants as being exposed to a 'complex' food web, whereas galls on treatment plants were exposed to a 'simple' food web.
In late August, we collected marked and bagged galls from each plant, placed them into 30 mL vials and <!-- allowed them to complete development--> kept them in the lab for 4 months at room temperature. 
We then opened galls under a dissecting scope and determined whether larva survived to pupation (our measure of fitness) or were parasitized. Since we were interested in selection imposed by interactions with parasitoids, we restricted our data to larva that either survived to pupation, was parasitized by an egg parasitoid (*Platygaster* sp.), or was parasitized by a larval parasitoid. For the food-web treatment that excluded parasitoids, we further restricted our data by removing any instances of parasitism by a larval parasitoid. This represented less than 3\% of the observations in this food-web treatment and allowed us to focus our inferences of selection on those imposed by the egg parasitoid.  
Together, we had survival estimates for 1,306 larva from 607 galls, 111 plants, and 8 plant genotypes.

@@

We collected data on three different traits that we anticipated would experience selection based on our previous work (@Barbour2016) and others work with Cecidomyiid midges (@Weis1983, @Heath2018). 
First, we measured gall diameter as the size of each gall chamber to the nearest 0.01 mm at its maximum diameter (perpendicular to the direction of plant tissue growth). 
Our previous work has shown that a larger gall diameter provides a refuge for larva from parasitoid attack (@Barbour2016). <!-- Consider whether it is important to make the distinction that diameter was measured at the multi-chambered gall level in previous work -->
Second, we measured the clutch size of adult female midges by counting the number of chambers in each gall (@Weis1983). 
All larva collected from the same multi-chambered gall were scored with the same clutch size. 
Third, we measured female preference for oviposition (egg-laying) sites as the density of larva observed on a plant in an independent survey. Specifically, we randomly sampled five branches per tree and summed the number of individual gall chambers observed. We then converted these counts to a measure of gall density per 100 shoots by counting the number of shoots on the last branch we sampled. All larva collected from the same plant were scored with the same female preference.
The measurement of larval densities on plants in the field is a commonly used index for measuring oviposition preference (@Gripenberg2010); however, caution must be taken in inferring 'preference' as larval densities can be influenced by processes other than preference (@Singer1986). Fortunately, a couple of features of our study system suggest that larval density on a plant may be a good proxy for female preference. For example, since our data comes from a randomized placement of willow genotypes in a common garden, there is no consistent bias in which willow genotypes that females are exposed to while searching for oviposition sites. Also, egg predation is a minor source of mortality for galling insects in general (@Hawkins1997), thus we do not expect any prior egg predation to bias our estimates of observed larval densities.

@@

To characterize the shape of the fitness landscape in simple and complex food webs, we first used a generalized linear mixed model to quantify selection surfaces on individual traits. We used a binomial error distribution (logit link function) since larval survival (0 or 1) was our response variable and measure of fitness. We specified linear and quadratic terms for each gall trait as well as linear interaction terms between each gall trait as fixed effects in the statistical models. To account for the correlated structure of clutch size (gall level) and female preference (plant level) as well as any independent effects of willow genotype on larval survival, we specified gall ID nested within plant ID nested within plant genotype as random effects. Since we were interested in characterizing the fitness landscape -- the relationship between mean trait values and population mean fitness -- we assumed the mean value of our random effects (i.e. setting them to zero) to estimate selection gradients. Also, the fitness landscape assumes that traits distributions are multivariate normal. To better meet this assumption, we log-transformed clutch size and added a small constant (1) to female preference before log transforming, since our surveys occassionaly estimated zero larval densities. We then scaled all phenotypic traits to mean=0 and SD=1 in order to calculate standardized selection gradients that were comparable across traits and with other studies of natural selection. We used the method of @Janzen1998 to calculate directional ($\beta_{z_i}$), quadratic ($\gamma_{z_i,z_i}$), and correlational ($\gamma_{z_i,z_j}$) selection gradients and used parametric bootstrapping (1000 replicates) to calculate their 95\% confidence intervals (@Bolker2009). We estimated directional selection gradients by excluding quadratic terms and statistical interactions in the model. Note that for visualizing the fitness landscape we restrict trait axes to $\pm 1$ SD of the mean trait value as this contains the majority of the trait distribution that selection is acting on.

@@

@@

@@

@@

\begin{thebibliography}{}

\bibitem[{Cook et~al.(2015)Cook, Collaborator, and Expert}]{CookEtAl2015}
Cook, O.~E., G.~H. Collaborator, and A.~Q. Expert. 2015.
\newblock Data from: Template and guidelines for using \LaTeX{} in \textit{The American Naturalist}.
\newblock American Naturalist, Dryad Digital Repository, http://dx.doi.org/10.5061/dryad.XYZAB.

\bibitem[{Davis et~al.(2011)Davis, Brakora, and Lee}]{DavisEtAl2011}
Davis, E.~B., K.~A. Brakora, and A.~H. Lee. 2011.
\newblock Evolution of ruminant headgear: a review.
\newblock Proceedings of the Royal Society B 278:2857--2865.

\bibitem[{Inglis et~al.(2011)Inglis, Roberts, Gardner, and Buckling}]{Ing11}
Inglis, R.~F., P.~G. Roberts, A.~Gardner, and A.~Buckling. 2011.
\newblock Spite and the scale of competition in \textit{Pseudomonas
  aeruginosa}.
\newblock American Naturalist 178:276--285.

\bibitem[{Lemod\`{e}le et~al.(2007)Lemodele, Kapitelschreiber, and Exemplar}]{LemKapEx07}
Lemod\`{e}le, P.-Q., A.~B. Kapitelschreiber, and C.~D.~E. Exemplar. 2007.
\newblock An exemplary instance of chapters in books.
\newblock Pages 231--245 \emph{in} J.-P. \'{E}crivain and M.~A. Term\'{e}szettud\'{o}s, eds. Inspiring Instances of Brilliant Writing. Truth Pudding Press, Fond du Lac, WI.

\bibitem[{Xiao et~al.(2015)Xiao, McGlinn, and White}]{Xiao2015}
Xiao, X., D.~J. McGlinn, and E.~P. White. 2015.
\newblock A strong test of the maximum entropy theory of ecology.
\newblock American Naturalist 185:E705--E80.

\end{thebibliography}

@@

\begin{table}[h]
\caption{Table for \textit{Iteomyia} Selection Gradients}
\label{Table:Gradients}
\centering
\begin{tabular}{lcc}
                                                                                  \\ \hline
                                  & \multicolumn{2}{c}{Food web (Estimate [95\% CI])} \\
\textbf{Selection gradient}       & \textbf{Complex} & \textbf{Simple}            \\ \hline
$\beta_{Diam}$                    & 0.47 [0.36-0.58] & 0.29 [0.23,0.37] \\
$\beta_{Clutch}$                  & 0.06 [-0.04,0.17] & -0.13 [-0.25,-0.04] \\
$\beta_{Pref}$                    & -0.24 [-0.36,-0.11] & -0.15 [-0.23,-0.07] \\
$\gamma_{Diam,Diam}$              & 0.13 [0.00,0.26] & 0.08 [0.00,0.16] \\
$\gamma_{Clutch,Clutch}$          & -0.10 [-0.25,0.04] & -0.06 [-0.16,0.02] \\
$\gamma_{Pref,Pref}$              & 0.39 [0.02,0.81] & -0.32 [-0.63,-0.03] \\
$\gamma_{Diam,Clutch}$            & -0.08 [-0.16,0.01] & -0.05 [-0.10,0.01] \\
$\gamma_{Diam,Pref}$              & -0.08 [-0.19,0.02] & -0.05 [-0.12,0.01] \\
$\gamma_{Clutch,Pref}$            & 0.04 [-0.06,0.13] & 0.02 [-0.04,0.08]            \\ \hline
\end{tabular}
\bigskip{}
\\
{\footnotesize Note: Table titles should be short. Further details should go in a `notes' area after the tabular environment, like this. $^a$ Okapis are not native to Chicago, but they are to be met with in both of the major Chicagoland zoos.}
\end{table}

>>>>

\documentclass[11pt]{article}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

% Added packages
\usepackage{amsmath} % for matrices
\usepackage{graphicx} % for organizing figures

@@

% I. Adaptive landscape is a powerful framework for understanding the evolution of biological diversity (or phenotypic evolution).

The adaptive landscape provides a powerful framework for understanding and predicting the evolution of biological diversity \textemdash from genes, to phenotypes, to species (Wright and Simpson). More than a metaphor, the adaptive landscape links quantitative genetic and phenotypic variation to evolution by natural selection (Lande 1979, Arnold papers). In particular, the shape of the adaptive landscape represents the presence and strength of different selection pressures imposed by the environment (Arnold's work). Species interactions often play a key role in defining the selective environment, as evidenced by the influence of resource competition (Schluter 2000), mutualisms (Jordano 1987), and predation (Abrams 2000) on the evolution of biological diversity. Although there is clear evidence that pairwise interactions shape the adaptive landscape, we also know that most species interact with multiple species in an ecological community. Understanding how the adaptive landscape is shaped by community context represents a major challenge for both empiricists and theoreticians given the diversity of selection pressures that a population is likely to experience in a natural ecological community.

% II. Species-interaction networks provide an explicit representation of the community context that populations are embedded within.

Species-interaction networks, such as a food web describing who eats whom, provide an explicit representation of the community context. These networks describe the interdependency of species within an ecological community, providing an effective framework for predicting how changes in the density of directly and indirectly connected species will impact other species, and in turn, community dynamics. Despite the importance of species-interaction networks for predicting ecological dynamics, these networks often lack quantitative information on how species interactions influence natural selection. Consequently, it is unclear how species-interaction networks shape the adaptive landscape, creating a major hurdle in our current ability to predict how changes in community context will affect phenotypic evolution.

% III. Here we combine these frameworks to predict how changes in community context will affect phenotypic evolution. (empirically?)

Here, we integrate species-interaction networks and the adaptive landscape to predict how changes in network structure will affect phenotypic evolution. We did this by first conducting a field experiment to characterize the effect of network structure on the adaptive landscape of a constituent species. We then used a mathematical model to simulate how changes in network structure altered phenotypic evolution. Specifically, we conducted a field experiment that manipulated the diversity of insect parasitoids that were able to impose selection on an abundant insect herbivore (\textit{Iteomyia salicisverruca})(Fig. 1). The larva of this herbivore species induce tooth-shaped galls when they feed on the developing leaves of willow trees (\textit{Salix} sp., @Russo2006). These galls provide protection from generalist predators (e.g. ants, spiders), thus the network of interacting parasitoids provides a realistic representation of the biotic environment this insect herbivore is experiencing. Therefore, our manipulation of parasitoid diversity alters the diversity of interactions, or food-web complexity, that this insect herbivore experiences.

% IV. Lay out predictions for how changes in network structure influence the adaptive landscape.

@@

% Confronting this challenge is important now more than ever before, given the rapid changes we are witnessing in the diversity and composition of local communities (cite global change papers).

% Natural selection is inherently an ecological process, describing how the environment changes phenotypic distributions of a population due to differences in fitness. %within a generation

%At the same time, evolutionary biologists have long recognized that changes in the biotic environment can alter the dynamics of natural selection. However, the biotic environment in which populations are evolving often remains a bit of a "black box" that's labelled by a general ecological process such as competition, predation, or mutualism. Because of this, it remains difficult to predict how changes in the biotic environment will affect the direction and magnitude of natural selection. Such predictions are urgently needed given the rapid changes in the biotic environment that most populations are currently experiencing throughout the world.

% For example, the Lande equation ($\delta z=G\beta$) predicts the evolutionary trajectory of phenotypes ($\delta z$) as a function of directional selection gradients ($\beta$) and the additive genetic variance and covariance between these phenotypes (G-matrix, $G$). Similarly, the curvature of the adaptive landscape ($\gamma - \beta \beta^\text{T}$). also shapes the G-matrix, which ultimately determines the ... . Importantly, the components of this theoretical framework can be est quantities that predict evolutionary change can all be estimated empirically

% People have often studied pairwise species interactions. There is also tantalizing evidence that community context can alter natural selection on phenotypic distributions, but predicting these changes remains difficult. This is because work hasn't explicitly embodied the environment (which a network can).

% Key part here is environment, which a food web or species-interaction network can provide an explicit representation of (i.e. community context).

% Biological diversity \textemdash from genes, to phenotypes, to species \textemdash has and continues to be shaped by the interplay between ecological and evolutionary processes. Much of this biological diversity has been molded by natural selection arising from species interactions, such as resource competition (Schluter 2000), mutualisms (Jordano 1987), and predation (Abrams 2000). While there is clear evidence that pairwise interactions can drive evolution, we also know that most species interact with multiple species in an ecological community. Understanding and predicting how evolutionary dynamics unfold in a community context represents a major challenge both empirically and theoretically, given the diversity of selection pressures that a population is likely to experience in an ecological community. Answering this question also has important applications, especially given the rapid changes we are witnessing in the diversity and composition of local communities (cite global change papers).

% is a key challenge and predicting how changes in ecological communities is challenging and eminently theoretical (@McPeek2017, @De_Mazancourt2008, @Guimaraes2017, @Nuismer2013).

%<!-- Do I need to give more credit here to work by @TerHorst2018 and others?). I think it would be useful to show that there is some field and lab experiments demonstrating that community context can alter evolutionary trajectories -->
%<!-- Given the rapid loss of species diversity we are experiencing throughout the world (cite), we are in urgent need of work that makes and tests predictions for how the loss of species will affect the evolutionary process in natural communities.

%<!-- Theoretical models have begun to examine how the network structure of species interactions drives evolutionary change (Nuismer paper; Guimeras paper; Ecology Letters paper from a spanish guy...; @McPeek2017); however, we are currently lacking experimental tests of how changes in network structure will alter the dynamics of natural selection in the field. -->

% The fitness landscape provides a unifying framework for linking the ecology and evolution of populations (Lande 2007; McPeek 2017). The average fitness of a population is a common currency in ecology and evolution, but usually goes by different names in each field. Ecologists refer to it as per-capita population growth rate ($dN/Ndt$), whereas evolutionary biologists call it the natural log of population mean fitness ($ln(\bar W_N)$). In addition to having different names, ecologists and evolutionary biologists have typically focused on different processes that shape the fitness landscape. For example, population ecologists have long studied the effect of a population's density on its per-capita growth rate (i.e. density-dependence, CITE Foundational and current work). In contrast, evolutionary biologists have focused on how the mean trait value of a population influences its average fitness, as this describes the direction and magnitude of natural selection (CITE foundational and current work). Therefore, the fitness landscape describes the joint ecological and evolutionary dynamics of a population in a given environment.

% Community ecologists have extended the ecological side of the fitness landscape by incorporating network theory. Species-interaction networks, such as a food web describing who eats whom, provide an explicit representation of the biotic environment as they describe the interdependency of populations within an ecological community. This has provided an effective framework for predicting how changes in the biotic environment (e.g. density of directly and indirectly connected species) will impact population dynamics within species-rich communities. At the same time, evolutionary biologists have long recognized that changes in the biotic environment can alter the dynamics of natural selection. However, the biotic environment in which populations are evolving often remains a bit of a "black box" that's labelled by a general ecological process such as competition, predation, or mutualism. Because of this, it remains difficult to predict how changes in the biotic environment will affect the direction and magnitude of natural selection. Such predictions are urgently needed given the rapid changes in the biotic environment that most populations are currently experiencing throughout the world.

@@

We conducted our study within a four-year old common garden of coastal willow (\textit{Salix hookeriana}) located at Humboldt Bay National Wildlife Refuge (HBNWR) (40\textdegree 40'53"N, 124\textdegree 12'4"W) near Loleta, California, USA. This common garden consists of 26 different willow genotypes that were collected from a single population of willows growing around Humboldt Bay. Stem cuttings of each genotype (25 replicates per genotypes) were planted in a completely randomized design in two hectares of a former cattle pasture at HBNWR. Willows in our garden begin flowering in February and reach their peak growth in early August. During this study, willows had reached 5 - 9m in height. Further details on the genotyping and planting of the common garden are available in @Barbour2015.

@@

We setup our food-web manipulation across 128 plants soon after galls began developing on *S. hookeriana* in early June of 2013. 
These 128 plants came from eight different plant genotypes, spanning the range of trait variation observed in this willow population (@Barbour2015). 
On treatment plants (8 replicates per genotype), we enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species that attack during larva development (hereafter larval parasitoids). 
This treatment did not exclude the egg parasitoid *Platygaster* sp. which attacks prior to gall initiation (note that in Cecidomyiid midges, larva initiate gall development CITE). 
On control plants (8 replicates per genotype), we used flagging tape to mark 14 galled leaves per plant (~30 larva), allowing the full suite of parasitoids to attack \textit{Iteomyia}. 
Marking galls with flagging tape ensured that we compared control and treatment galls with similar phenology when we collected galls later in the season. 
Our food-web manipulation altered the average number of trophic interactions that \textit{Iteomyia} was exposed to from BLANK on control plants to BLANK on treatment plants. 
Thus, we refer to galls on control plants as being exposed to a 'complex' food web, whereas galls on treatment plants were exposed to a 'simple' food web.
In late August, we collected marked and bagged galls from each plant, placed them into 30 mL vials and kept them in the lab for 4 months at room temperature. 
We then opened galls under a dissecting scope and determined whether larva survived to pupation (our measure of fitness) or were parasitized. Since we were interested in selection imposed by interactions with parasitoids, we restricted our data to larva that either survived to pupation, was parasitized by an egg parasitoid (\textit{Platygaster} sp.), or was parasitized by a larval parasitoid. For the food-web treatment that excluded parasitoids, we further restricted our data by removing any instances of parasitism by a larval parasitoid. This represented less than 3\% of the observations in this food-web treatment and allowed us to focus our inferences of selection on those imposed by the egg parasitoid.  
Together, we had survival estimates for 1,306 larva from 607 galls, 111 plants, and 8 plant genotypes.

@@

We collected data on three different traits that we anticipated would experience selection based on our previous work (@Barbour2016) and others work with Cecidomyiid midges (@Weis1983, @Heath2018) (\citealt{Heath2018}). 
First, we measured gall diameter as the size of each gall chamber to the nearest 0.01 mm at its maximum diameter (perpendicular to the direction of plant tissue growth). 
Our previous work has shown that a larger gall diameter provides a refuge for larva from parasitoid attack (@Barbour2016). <!-- Consider whether it is important to make the distinction that diameter was measured at the multi-chambered gall level in previous work -->
Second, we measured the clutch size of adult female midges by counting the number of chambers in each gall (@Weis1983). 
All larva collected from the same multi-chambered gall were scored with the same clutch size. 
Third, we measured female preference for oviposition (egg-laying) sites as the density of larva observed on a plant in an independent survey. Specifically, we randomly sampled five branches per tree and summed the number of individual gall chambers observed. We then converted these counts to a measure of gall density per 100 shoots by counting the number of shoots on the last branch we sampled. All larva collected from the same plant were scored with the same female preference.
The measurement of larval densities on plants in the field is a commonly used index for measuring oviposition preference (@Gripenberg2010); however, caution must be taken in inferring 'preference' as larval densities can be influenced by processes other than preference (@Singer1986). Fortunately, a couple of features of our study system suggest that larval density on a plant may be a good proxy for female preference. For example, since our data comes from a randomized placement of willow genotypes in a common garden, there is no consistent bias in which willow genotypes that females are exposed to while searching for oviposition sites. Also, egg predation is a minor source of mortality for galling insects in general (@Hawkins1997), thus we do not expect any prior egg predation to bias our estimates of observed larval densities.

@@

To characterize the shape of the fitness landscape in simple and complex food webs, we first used a generalized linear mixed model to quantify selection surfaces on individual traits. We used a binomial error distribution (logit link function) since larval survival (0 or 1) was our response variable and measure of fitness. We specified linear and quadratic terms for each gall trait as well as linear interaction terms between each gall trait as fixed effects in the statistical models. To account for the correlated structure of clutch size (gall level) and female preference (plant level) as well as any independent effects of willow genotype on larval survival, we specified gall ID nested within plant ID nested within plant genotype as random effects. Since we were interested in characterizing the fitness landscape -- the relationship between mean trait values and population mean fitness -- we assumed the mean value of our random effects (i.e. setting them to zero) to estimate selection gradients. Also, the fitness landscape assumes that traits distributions are multivariate normal. To better meet this assumption, we log-transformed clutch size and added a small constant (1) to female preference before log transforming, since our surveys occassionaly estimated zero larval densities. We then scaled all phenotypic traits to mean=0 and SD=1 in order to calculate standardized selection gradients that were comparable across traits and with other studies of natural selection. We used the method of @Janzen1998 to calculate directional ($\beta_{z_i}$), quadratic ($\gamma_{z_i,z_i}$), and correlational ($\gamma_{z_i,z_j}$) selection gradients and used parametric bootstrapping (1000 replicates) to calculate their 95\% confidence intervals (@Bolker2009). We estimated directional selection gradients by excluding quadratic terms and statistical interactions in the model. Note that for visualizing the fitness landscape we restrict trait axes to $\pm 1$ SD of the mean trait value. This emphasizes the fact that we can only reliably estimate the shape of the adaptive landscape near the mean trait values of the population (Arnold 2001, Morrissey 2009).

@@

\subsection*{Simulating Phenotypic Evolution}

We used the method proposed by Hansen and Houle (2008) to characterize the evolvability of the G-matrices resulting from different adaptive landscapes. Their index of evolvability is theoretically linked to the theory underlying G matrices and adaptive landscapes, and describes the ability of a population to evolve in the direction of selection when stabilizing selection is absent.

% Hansen and Houle 2008
%Alternatively, but not
%inconsistently, evolvability can be viewed in a quantitative
%genetics context as a short-term ability to respond to
%directional selection (Houle, 1992), which depends on
%standing additive genetic variation in the traits under
%selection and on the autonomy of this genetic variation
%from other traits under stabilizing selection (Hansen
%et al., 2003a, b; Hansen, 2003; Hansen & Houle, 2004)

@@

\subsection*{Shape of the adaptive landscape}

The shape of the adaptive landscape is described by its slope ($\beta$) and curvature ($\gamma - \beta \beta^\text{T}$). We found that food-web complexity altered the slope of the adaptive landscape in two key ways. First, there were fewer traits under directional selection in the complex (2 of 3) vs. simple food web (3 of 3). For example, across both food-web treatments there was directional selection for larger gall chambers and for adult females to avoid ovipositing on trees with high densities of conspecifics (fig.~\ref{Fig:FL_1D}; table~\ref{Table:Gradients}). In the simple food web, we also observed directional selection for smaller clutch sizes, but there was no evidence of selection acting on this trait in the complex food web (fig.~\ref{Fig:FL_1D}; table~\ref{Table:Gradients}). The absence of selection on clutch size in the complex food web appeared to be a result of conflicting selection pressures imposed by each guild of parasitoids. For example, when we subset our data to focus on differences between parasitoid guilds, we found that larval parasitoids actually impose directional selection for larger clutch sizes (\beta_{diam}=0.12 [0.04,0.21]).

The second key effect of food-web complexity was to steepen the slope of the adaptive landscape (i.e. increase strength of selection). For example, the magnitude of directional selection in complex vs. simple food webs was ~60\% higher for both chamber diameter and female preference. This result was not due to any difference in the relationship between larva survival and gall phenotype (INSERT REGRESSION ANALYSIS), but was a result of the fact that average larval survival was lower in complex (INSERT MEAN PROBABILITY OF SURVIVAL) vs. simple food webs (INSERT MEAN PROBABILITY OF SURVIVAL). Remember that the selection gradient can be expressed as the covariance between absolute fitness ($W$) and individual phenotype ($z$) divided by mean fitness ($\bar W$). Therefore, lower survival, and thus mean fitness, will always increase the strength of selection.

Unlike the slope, the curvature of the adaptive landscape is influenced by directional ($\beta$), quadratic ($\gamma_{z_i,z_i}$), and correlational ($\gamma_{z_i,z_j}$) selection gradients. Our food-web treatment did not appear to alter correlational selection, and there was no strong evidence that any of these selection gradients differed from zero (table~\ref{Table:Gradients}). Similarly, our food-web treatment did not appear to influence quadratic selection on chamber diameter or clutch size, and there was no strong evidence that these selection gradients differed from zero (table~\ref{Table:Gradients}). In contrast, our food-web treatment did alter quadratic selection acting on female preference. Specifically, we found evidence for disruptive selection (selection for extreme phenotypes) in the complex food web, but no evidence of disruptive or stabilizing selection in the simple food web (fig.~\ref{Fig:FL_1D}; table~\ref{Table:Gradients}).

The curvature ($\gamma - \beta \beta^\text{T}$) of the adaptive landscape describes how natural selection changes genetic variation ($\Delta \text{G}$) within a generation.

In the simple food web, we found that the adaptive landscape acted to decrease additive genetic variance in all three traits (diagonals of matrix).

$$\text{Curvature}_{simple} = \begin{pmatrix} -0.05 & 0.02 & 0.03 \\  0.02 & -0.01 & -0.01 \\  0.03 & -0.01 & -0.02 \end{pmatrix}$$

In contrast, we found the adaptive landscape in the complex food web only decreased the additive genetic variance in one of the three traits (chamber diameter).

\text{Curvature}_{complex} = \begin{pmatrix} -0.13 & 0 & 0.9 \\ 0 & 0 & 0 \\ 0.09 & 0 & 0.33 \end{pmatrix}

In the complex food web, we found that the adaptive landscape  We found that the adaptive landscape complex food web,

We found that there were fewer selective constraints influence *Iteomyia* in the complex vs. the simple food-web treatment. Specifically, the curvature of the complex food-web was described as:

whereas the curvature of the simple food-web is summarized by the following matrix:

Specifically, we found that simple food webs would result in a decrease in additive genetic variance for all phenotypic traits (matrix diagonals) within a generation, wherease in the complex food web, we only found that one trait (chamber diameter), would experience a decrease in additive genetic variance (albeit large). Any non-zero for the off diagonals is evidence of constraints. In the simple food-web, we found evidence of constraints on each additive genetic covariance, albeit small, whereas for the complex food-web, we only observed a constraint on the covariance between chamber diameter and female preference (favoring trait intergration). Taken together, we observe constrains on 2 of 6 trait variances and covariances in the complex food web, whereas 6 of 6 trait variances and covariances were constrained in the simple food web.

Our two dimensional surfaces are likely representative given the fact that there is no strong evidence of correlational selection (Phillips and Arnold 1989, Evolution)

\subsection*{Phenotypic evolution is less constrained in the complex food web}

% of the adaptive landscape describe its shape and therefore the direction and magnitude of phenotypic ($\Delta \text{z}$) and genetic change ($\Delta \text{G}$).

Three key patterns emerged from our simulations of phenotypic evolution. The first was that phenotypic evolution was faster in complex vs. simple food webs. Specifically, it took \textit{Iteomyia} twice as long to hit the boundary of the adaptive landscape in the simple vs. complex food web. This pattern was driven by the fact that the absolute strength of selection was larger in complex vs. simple food webs, as we no longer observed this strong difference if we standardized all selection gradients by average larval survival measured in the complex food web (Appendix).

The second key pattern we observed was that phenotypic evolution was less constrained in complex compared to simple food webs.

Finally, we found that the adaptive landscape in the complex food web enhanced the evolvability of \textit{Iteomyia} compared to the simple food web.

Food-web complexity fundamentally altered the shape of the adaptive landscape. For example,

Across both food-web treatments, there was no clear evidence of correlational selection between (Table 1).

We observed disruptive selection on female preference in the complex, but no evidence of any nonlinear selection the simple food web. This disruptive selection was likely driven by a different functional response of larval parasitoids to female preference. For example, when we subset our data to focus on differences between larval and egg parasitoids, we found that there was a trend for larval parasitoids to impose disruptive selection (\gamma_{pref,pref}=0.22 [-0.06,0.49]).

In both complex and simple food webs, chamber diameter was under strong directional selection with larger galls resulting in higher larval survival (fig.~\ref{Fig:FL_1D}; table~\ref{Table:Gradients}).

In general, if there was evidence of selection, then the magnitude of selection was always greater in complex vs. simple food webs. These steeper selection gradients were simply due to the fact that average larval survival was lower in complex food webs.

@@

Our key finding was that phenotypic evolution was less constrained for a species that was embedded within a complex food web compared to a simple one. These

Two key patterns emerged from our study. First, selection was stronger and thus the potential speed of evolution in complex compared to simple food webs. Second, the adaptive landscape was less constrained in the complex vs. the simple food web.

@@

\bibliographystyle{amnat}
\bibliography{bibliography}

@@

\begin{table}[h]
\caption{Table for \textit{Iteomyia} Selection Gradients}
\label{Table:Gradients}
\centering
\begin{tabular}{lcc}
                                                                                  \\ \hline
                                  & \multicolumn{2}{c}{Food web (Estimate [95\% CI])} \\
\textbf{Selection gradient}       & \textbf{Complex} & \textbf{Simple}            \\ \hline
$\beta_{Diam}$                    & 0.47 [0.36,0.59] & 0.29 [0.23,0.37] \\
$\beta_{Clutch}$                  & 0.06 [-0.05,0.19] & -0.09 [-0.17,-0.02] \\
$\beta_{Pref}$                    & -0.24 [-0.38,-0.11] & -0.15 [-0.24,-0.07] \\
$\gamma_{Diam,Diam}$              & 0.13 [0.00,0.26] & 0.08 [0.00,0.16] \\
$\gamma_{Clutch,Clutch}$          & -0.10 [-0.25,0.05] & -0.06 [-0.16,0.03] \\
$\gamma_{Pref,Pref}$              & 0.39 [0.04,0.74] & -0.07 [-0.25,0.11] \\
$\gamma_{Diam,Clutch}$            & -0.08 [-0.16,0.01] & -0.05 [-0.10,0.00] \\
$\gamma_{Diam,Pref}$              & -0.08 [-0.18,0.01] & -0.05 [-0.11,0.01] \\
$\gamma_{Clutch,Pref}$            & 0.04 [-0.05,0.13] & 0.02 [-0.03,0.08]            \\ \hline
\end{tabular}
\bigskip{}
\\
{\footnotesize Note: Table titles should be short. Further details should go in a `notes' area after the tabular environment, like this. $^a$ Okapis are not native to Chicago, but they are to be met with in both of the major Chicagoland zoos.}
\end{table}
